---
title: "Activity - Z Scores"
author: "Willis/Wells"
date: "2025-11-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


 "Baltimore County 911 Overdose Calls Analysis"
author: Benjamin Messinger


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Z-Score Analysis of Baltimore County 911 Calls

## Introduction

In this analysis, we'll examine data on 911 drug overdose calls from Baltimore County to identify unusually busy or quiet days using z-scores. By calculating the number of calls per day and analyzing their statistical distribution, we can determine which days experienced an abnormal volume of emergency calls - information that could lead to potential story ideas.

A z-score measures how many standard deviations a data point is from the mean. Generally:
- Z-scores between -2 and 2 represent typical values (about 95% of the data)
- Z-scores beyond ±2 indicate unusual values that warrant investigation
- Z-scores beyond ±3 are very unusual (only about 0.3% of normally distributed data)

## Task 1: Setup and Data Loading

Let's start by loading the necessary packages and our dataset:

```{r load-packages}
library(tidyverse)  # For data manipulation and visualization
library(lubridate)  # For handling dates
library(knitr)      # For nice tables
library(scales)     # For better axis formatting
```

```{r load-data}
# Load the 911 calls data
calls_data <- read_csv("https://raw.githubusercontent.com/dwillis/datajournalismbook-maryland/refs/heads/main/data/baltco_911_calls.csv")

# Display the first few rows to understand the structure
head(calls_data)
```

## Task 2: Initial Data Exploration

Let's explore the basic structure and characteristics of our dataset:

```{r explore-data}
# Get a summary of the data
glimpse(calls_data)

# Check for any missing values
colSums(is.na(calls_data))

# Check the date range in our data
min_date <- min(calls_data$date)
max_date <- max(calls_data$date)
cat("Data ranges from", min_date, "to", max_date)

```

## Task 3: Aggregating Calls by Day (3 points)

Now let's count how many 911 overdose calls occurred on each day:

```{r aggregate-by-day}
# Count calls per day
daily_calls <- calls_data |>
  count(date, name = "call_count")

# Display the first few days
head(daily_calls, 10) 

# Basic summary statistics for daily call volume
summary(daily_calls$call_count)
```

Let's visualize the time series of daily calls. What patterns do you see? 

**Write your answer here**

The maximum amount of calls was 23 on Oct. 4, 2022 and the minimum amount of calls was one call on Jun. 11 and Feb. 6. The high-end of the amount of calls that occur toward the begining of the month. That said, there are really no true patterns, the median of 10 calls per day and mean of 10.44 is accurate to the majority of the data in the set. The data fluxuates throughout the high and low ends of the data. 


```{r daily-calls-plot}
# Plot daily call volume
ggplot(daily_calls, aes(x = date, y = call_count)) +
  geom_line(color = "darkblue", alpha = 0.7) +
  geom_point(color = "darkblue", alpha = 0.5, size = 1) +
  theme_minimal() +
  labs(title = "Daily 911 Overdose Calls in Baltimore County",
       x = "Date",
       y = "Number of Calls",
       caption = "Source: Baltimore County 911 Call Data") +
  scale_y_continuous(labels = comma) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```




## Task 4: Calculating Z-Scores for Daily Call Volume

Now, let's calculate z-scores to identify statistically unusual days:

```{r calculate-z-scores}
# Calculate z-scores for daily call counts
daily_calls <- daily_calls |>
  mutate(
    mean_calls = mean(call_count),
    sd_calls = sd(call_count),
    z_score = (call_count - mean_calls) / sd_calls
  )

# Display the data with z-scores
daily_calls |>
  select(date, call_count, z_score) |>
  head(10) 
```

## Task 5: Identifying Unusual Days (3 points)

Let's identify days with unusually high or low call volumes (|z-score| > 2):

```{r identify-unusual-days}
# Find days with unusual call volumes
unusual_days <- daily_calls |>
  filter(abs(z_score) > 2) |>
  arrange(desc(abs(z_score)))

# Display unusual days
unusual_days |>
  select(date, call_count, z_score) |>
  kable(digits = 2)

# Count how many days were unusually busy vs. unusually quiet
unusual_days |>
  group_by(unusual_type = if_else(z_score > 0, "Unusually High", "Unusually Low")) |>
  summarise(count = n())
```

Let's visualize these unusual days in the context of the full dataset. How do they differ from your answer on Task 3?

My answer on Task 3 didn't truly recognize the difference between there being more high days than low days, that said looking at the graph it makes sense that there's more high spikes than low spokes. But, the idea that there's a threshold of a day could be unusually high or low is not present in the data. Is the high end of being unusual is 18 and not all eight days that have 17 calls. Also, why is three the threshold for it being unusually low? So yes, it does differ from my answer because I didn't have set thresholds of a day being unusual.


```{r plot-unusual-days}
# Create a clean factor for the plot legend
daily_calls <- daily_calls |>
  mutate(call_volume_type = case_when(
    z_score > 2 ~ "Unusually High",
    z_score < -2 ~ "Unusually Low",
    TRUE ~ "Normal"
  ))

# Plot with unusual days highlighted
ggplot(daily_calls, aes(x = date, y = call_count, color = call_volume_type)) +
  geom_line(color = "gray70") +
  geom_point(size = 2) +
  scale_color_manual(values = c("Normal" = "gray50", 
                               "Unusually High" = "red", 
                               "Unusually Low" = "blue")) +
  theme_minimal() +
  labs(title = "Daily 911 Overdose Calls with Unusual Days Highlighted",
       subtitle = "Unusual days have z-scores with absolute values > 2",
       x = "Date",
       y = "Number of Calls",
       color = "Call Volume",
       caption = "Source: Baltimore County 911 Call Data") +
  scale_y_continuous(labels = comma) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Task 6: Day of Week Patterns

Let's also check if certain days of the week typically see more 911 calls:

```{r day-of-week-analysis}
# Add day of week to our dataset
daily_calls <- daily_calls |>
  mutate(
    day_of_week = wday(date, label = TRUE)
  )

# Calculate average calls by day of week
#dow = Day of Week
dow_summary <- daily_calls |>
  group_by(day_of_week) |>
  summarise(
    avg_calls = mean(call_count),
    median_calls = median(call_count),
    min_calls = min(call_count),
    max_calls = max(call_count)
  ) |>
  arrange(match(day_of_week, c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Display summary by day of week
dow_summary |> kable(digits = 1)

# Plot calls by day of week
ggplot(dow_summary, aes(x = day_of_week, y = avg_calls)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = round(avg_calls, 1)), vjust = -0.5) +
  theme_minimal() +
  labs(title = "Average 911 Overdose Calls by Day of Week",
       x = "Day of Week",
       y = "Average Number of Calls",
       caption = "Source: Baltimore County 911 Call Data") +
  theme(axis.text.x = element_text(angle = 0))
```

## Task 7: Accounting for Day of Week Pattern (3 points)

We can also calculate z-scores that account for the day of the week pattern. How would you describe the day with the highest z-score without using the term "z-score"?

**Write your answer here**

Weekends have higher amounts of calls than any other day of the week, especially Fridays and Saturdays, but also Sundays. Saturdays saw the highest number of calls, compared to Friday having the second-most and Sunday had the third-most. 

```{r day-of-week-z-scores}
# Calculate day-of-week adjusted z-scores
daily_calls <- daily_calls |>
  group_by(day_of_week) |>
  mutate(
    dow_mean = mean(call_count),
    dow_sd = sd(call_count),
    dow_z_score = (call_count - dow_mean) / dow_sd
  ) |>
  ungroup()

# Find days that were unusual for their day of the week
dow_unusual_days <- daily_calls |>
  filter(abs(dow_z_score) > 2) |>
  arrange(desc(abs(dow_z_score)))

# Display days that were unusual for their day of the week
dow_unusual_days |>
  select(date, day_of_week, call_count, dow_mean, dow_z_score) 
```


## Task 8: Temporal Patterns and Trends

Let's also look for seasonal patterns or trends over time:

```{r temporal-analysis}
# Add month and year for seasonal analysis
daily_calls <- daily_calls |>
  mutate(
    month = month(date, label = TRUE),
    year = year(date)
  )

# Calculate average calls by month
monthly_summary <- daily_calls |>
  group_by(month) |>
  summarise(
    avg_calls = mean(call_count),
    median_calls = median(call_count)
  )

# Plot calls by month
ggplot(monthly_summary, aes(x = month, y = avg_calls, group = 1)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = "Average 911 Overdose Calls by Month",
       x = "Month",
       y = "Average Number of Calls",
       caption = "Source: Baltimore County 911 Call Data")

# Show a rolling 7-day average to better see trends
daily_calls <- daily_calls |>
  arrange(date) |>
  mutate(rolling_avg_7day = zoo::rollmean(call_count, k = 7, fill = NA))

# Plot with 7-day rolling average
ggplot(daily_calls, aes(x = date)) +
  geom_line(aes(y = call_count), color = "gray60", alpha = 0.7) +
  geom_line(aes(y = rolling_avg_7day), color = "blue", size = 1) +
  theme_minimal() +
  labs(title = "Daily 911 Overdose Calls with 7-Day Moving Average",
       x = "Date",
       y = "Number of Calls",
       caption = "Source: Baltimore County 911 Call Data") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Task 9: T Most Unusual Days

For the most unusual days, we might want to check if they coincide with significant events:

```{r investigate-top-outliers}
# Get the top 10 most unusual days
top_outliers <- daily_calls |>
  arrange(desc(abs(z_score))) |>
  head(10) |>
  select(date, day_of_week, call_count, z_score)

# Display these days
top_outliers 
```

## Task 10: Journalistic Insights (6 points)

Based on our z-score analysis, what are some potential angles for reporting? What would your next steps be and what would you tell an editor about your findings?

**Write your answer here**


The angles I would take in this z-score analysis is reporting why these days specifically had more or less calls. Jan. 1 makes the most sense because it's New Years Day, but why Oct. 4, Jul. 14 and Jul. 22. Also why did Feb. 6, Jun. 11 and Dec. 21 had the least amount of calls, was there nothing going on on those days. My next steps would be researching if there were events happening in those days and I would say that there are trends that factor in to there being more or less calls to 911. I would tell my editor that there is some type of reason that there are more Overdose related 911 calls on some days than others.
