---
title: "Oct 1 Activities: M&Ms; Survey; Sampling"
date: 9-30-2025
---
Benjamin Messinger

All three class activities are on this .rmd notebook

Activity #1: Activity: M&Ms - Central Limit Theorem

Working in groups, each group will get a handful of candy from the bag, and then count the number of red and black candies. Log those in this Google Sheet 

https://docs.google.com/spreadsheets/d/1KOeoVTq2W7L2T9sWjwh5U_2f59GsTeAylVGD53rNWww/edit?usp=sharing

(give your group a nice name). 7 of a kind

Take a look at the values that we all get - what do you notice about the patterns? This is an exercise about the central limit theorem. 

The central limit theorem states that the sampling distribution of the mean of any independent, random variable will be normal or nearly normal, if the sample size is large enough.

Then discuss and try to answer these questions:

--How close is your group's mean to the true proportion of 0.5?
--Why do different groups get different results?
--What happens to individual extreme results when we combine all groups?
--How does this relate to polls and surveys in real life?

Write your answer below. And then you may eat the M&Ms.


**A:**
Q1: Out group's mean is 90% to the true proportion of 0.5
Q2: Different groups may get different results as other groups could have different proportions as they come out of the bag.
Q3: If we were to combine all groups, the groups would balance out and would be closer to normal.
Q4: This relate's to polls in real life because if you add up all test groups or polled people to all the peope who voted the results will be the true proportion. If the total amount of M&M's are an entire population and the colors are different survey options. 

------------------------------------------------------------------------------

# Activity #2:

Belonging Survey

Working in pairs, take a look back at the main findings from UMD's Belonging Survey:

https://president.umd.edu/internal/belonging


Specifically, focus on the "Sample Description" section and the Appendix. Discuss the differences and between the population and the sample, and what the document says about how the survey attempted to deal with them. Are there any decisions described in the document that raise questions in your minds? Detail the differences you see and your understanding of the methods they employed (it's ok if you don't understand all of them), and add your questions. You should discuss this as a pair but submit individually.

**A:**
The first question I have is the fact that 23% of the entire population of the 12,233 population responded and if the background of the people who responded to this sample creates bias of different groups are more likely to respond to the survey rather than others. The appendix does split up undergraduates, graduates, race and gender, but despite they look straight forward, there are different percentages.  
------------------------------------------------------------------------------

# TASK 3: SAMPLING

Based on the class sampling demo (available on Elms), sample data from a population using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

**A**
```{r}
library(tidyverse)


```


### Task 2: Load 1,000 rows of data with the ages of people from 18-80

**A**
```{r}
set.seed(42)
younguns <- tibble(
  age = sample(18:80, 1000, replace = TRUE)
)

head(younguns)

```

### Task 3: Sample 100 rows of data from the dataset
**A**
```{r}
sample_100 <- younguns |>
sample_n(100)

```


### Task 4: Calculate the mean age and standard deviation of the sample
**A**
```{r}
sample_100 |> summarize(mean = mean(age), median = median(age), sd =sd(age))
```



### Task 5: Calculate the mean age and standard deviation of the population
**A**
```{r}

younguns |> summarize(mean = mean(age), median = median(age), sd =sd(age))
```


### Task 6: Compare the sample and population statistics
**A**
```{r}


```

Compare the means and standard deviations. Are they similar? Go back and re-run the sampling and statistics calculations with a smaller sample size. How do the results change?

**A**: The mean and median of both data sets are similar, but are higher in the real data than the sample data. That said, the standard deviation in the sample data is higher than in the real data. If I went back and reran the sample calculations with a smaller sample size there would be a greater difference between the two samples. 


When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
