---
title: "Data Distribution, Volleyball Percentiles, Car Thefts, "
author: "Derek Willis /Rob Wells"
date: 09-23-2025
---

This workbook contains three exercises for our class. Complete them and hand it in at the end of class via a GitHub link.

Your name:
Benjamin Messinger
#------------------------------------------------------------------------------#


# Part 1: When Data Isn't Normal

Working in pairs, consider the following question: "Why do many statistical methods assume a normal distribution, and what challenges arise when data does not meet this assumption?"

First, come up with two examples of data that might not be normally distributed - in other words, that could be skewed or unequal in some way - and explain why they wouldn't be normally distributed. Feel free to draw what you think the distribution would look like.

Then, how should a reporter working on a story about data that isn't normally distributed describe that data? Which statistical measures that we've covered in class should they use or avoid when doing so? Use the examples you came up with, and you should discuss your answers together, but each student submits separately.

**Your answer**
If you're comparing two different sports teams or players and they have played different amounts of games, meaning that a team's amount of wins or a players amount of goals wouldn't be destributed normally. You would have to create percentages to calculate wins per games or goals per games to compare which player or team is better. 

Another example is if sales of seasonal fruits throughout a year, if a fruit that is only really grown in the spring and summer months, it would be obviously more likely that those months will have more sales throughout those months than other times of the year. You would need to compare it to that time of year from the previous years rather than the other times of that year.  



#------------------------------------------------------------------------------#


# Part 2: Volleyball Percentiles

We'll be calculating percentiles to see how good Maryland was in service aces in the 2024 season. We'll use `mutate()` and a new function `ntile()` to calculate percentiles. Get started by loading the tidyverse in a code block, then load the CSV file here: https://raw.githubusercontent.com/dwillis/NCAAWomensVolleyballData/refs/heads/main/data/ncaa_womens_volleyball_matchstats_2024.csv" and save it to a dataframe called `matches`.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the data

```{r}
matches <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensVolleyballData/refs/heads/main/data/ncaa_womens_volleyball_matchstats_2024.csv")
```

### Task 3: Create totals of aces for each team using `group_by()` and `summarize()`

```{r}
team_aces <- matches |> 
  group_by(team) |> 
  summarize(total_aces = sum(aces)) |> 
  arrange(desc(total_aces))
```


### Task 4: Calculate percentiles using `mutate()` and `ntile()`

```{r}
team_aces |> mutate(percentile = ntile(total_aces, 100))
```


### Task 5: Describe the percentiles

Write a couple of sentences that puts Maryland into context with the other teams in the dataset. You can do some Internet research to help with this.

**Answer**:
According to the data, Maryland volleyball sits in the 95th percentile for aces in its season. Meaning that the Terps have the 17th most aces in NCAA volleyball and have more than 95% of teams in the nation with 223 aces. UTEP is in the 100th percentile because it has the most aces in the nation with 263 aces. Maryland is in the same percentile as Creighton and Merrimack and have 15 teams ahead of the Terps in terms of percentile, despite having a lesser amount of aces. 


#------------------------------------------------------------------------------#

# Part 3: Car Thefts

Follow along with the demo to learn how to calculate standard deviation and how to create a histogram of the data using R and the Tidyverse. Get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)

```


### Task 2: Load the car thefts dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_jan24.csv


```{r}
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_jan24.csv")
```

### Task 3: Calculate the standard deviation of the number of car thefts, along with the mean

```{r}
car_thefts |> summarize(mean = mean(count), sd = sd(count))
```


### Task 4: Create a histogram of the number of car thefts


```{r}
car_thefts |> 
  ggplot() + 
  geom_histogram(aes(x = count), binwidth = 2)
```




### Task 5: Add a line to the histogram that shows the mean of the number of car thefts, and two other lines to show one standard deviation in both directions

```{r}
car_thefts |> 
  ggplot() + 
  geom_histogram(aes(x = count), binwidth = 2) +
  geom_vline(aes(xintercept = mean(count)), color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(count) - sd(count)), color = "blue", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(count) + sd(count)), color = "blue", linetype = "dashed", size = 1)
```

### Task 6: Describe your results

Is the distribution normal or skewed in one direction? Write a sentence in which you describe the range of the data covered by one standard deviation from the mean on both sides - how much of the data falls in that range?


**Answer**:
The distribution is slightly skewed to the right, but it still is contained in one standard deviation, meaning the majority of the data is present in the two right standard deviations, but mostly compact within the third standard deviation, which is about 60-75% of the standard deviation of the data. 




When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.

Congrats!


