---
title: "HW2 Prince George's County Crime"
---
Ben Messinger

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
install.packages(tidyverse)
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?

ANSWER HERE
Accidents increased throughout the year, first decreasing in february and march and then increasing to its highests points in May and October. The highest data point in the set was accidents in October at 569. Auto theft decreased then increased and then decreased again throughout the year. The most intriguing categories are certainly the ones relating to Auto theft, auto stolen and auto stolen & recovered. I think that all three of those categories show a direct relationship between the trends of crimes relating to cars in Prince Georges County. 

### Task 4: Calculating Averages

Calculate the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.

```{r}
pg_crime_averages <- pg_crime |> 
  mutate(
    total_incidents = rowSums(across(Jan:Dec), na.rm = TRUE),
    average_incidents = rowMeans(across(Jan:Dec), na.rm = TRUE)
  )



```
*I used AI on this task... It took me about 2 hours to complete and I couldn't figure it out, I tried 15 different ways to do it and it didn't work. I asked it to fix this code that I had for the dataset and code #total_incidents <- pg_crime |> mutate(total_incidents = rowSums(across(2:13))) pg_crime_averages <- pg_crime |> mutate(average_incidents = mean(rowSums(across(2:13))))

### Task 5: Describe the Averages

Examine the new dataframe, `pg_crime_averages`, by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the average number of incidents per month for each category. Which categories have the highest average number of incidents per month? Which have the lowest? What does this tell you about the pattern of crime in Prince George's County? Are there categories that you have questions about? Why or why not?

The highest amount of average incidents is Accidents, followed by Auto Stolen and Theft from Auto. The lowest amount of average incidents is residential robbery. I think it tells me that auto crimes are very common in Prince Georges county and it shows that the highest ammount of incidents involve some type of cars or automobiles. It also shows that ressidential robberies aren't very commmon, maybe because there are a ton of residences that can be robbed. 


### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
summer_crime_averages <- pg_crime_averages |> 
  mutate(
    june_change = ((Jun - average_incidents) / average_incidents) * 100,
    july_change = ((Jul - average_incidents) / average_incidents) * 100,
    august_change = ((Aug - average_incidents) / average_incidents) * 100
  )

```
*I also used AI with this code, I struggled with it for some time and couldn't figure out that I needed to subtract and divide it by average_incidents... So was trying to do it from Month to month and it wasn't working. 
### Task 7: Describe the Summer Crime Averages

Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

The percentage change for the categories involving automobiles were up in the summer months, for each month. Where every other type of incidents with the exception of homocides decreased during the summer. To me it tells me that because people are more inclined to be driving in the summer months and the weather is nicer, more vehicle-related crimes occur. I also noticed with several categories that the month to month changes even within the summer months would drastically change.
### Task 8: What's the Story?

Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Crimes relating to vehicles have skyrocketted in Prince Georges County. 

Auto-thefts and stolen cars had the two highest average amounts of incidents over any given month throughout the year. It has shown to become an epidemic in the county.

Moreover, in the summer where most incidents of crimes would decrease in some capacity between June, July and August, Auto-related crimes were consistently increased.  

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
