---
title: "HW3 WH Salaries"
---

Benjamin Messinger

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2024.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

The first thing that stands out to me is that there's 11 employees that have a salary of 0. I wonder if they are like unpaid interns or they actually have a salary that is not reported in the data set. Those 0 salaries will cause problems for analysis. The highest salaries were all detailies all making under $200,000 with the exception of Michelle Barrans who makes $250,000+, why does she make so much more than everyone else?

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |>
summarize(
  mean_salary = mean(salary, na.rm=TRUE),
 median_salary = median(salary, na.rm=TRUE))
```
Mean - $107,036.6	 | median - $110,500
### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |>
summarize(
  mean_salary = mean(salary, na.rm=TRUE),
 median_salary = median(salary, na.rm=TRUE),
max_salary = max(salary, na.rm=TRUE),
min_salary = min(salary, na.rm=TRUE))
```
Max - $251,258 Min - $0
### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The mean and median salaries show me that the average white house employee makes $7-10 thousand more than 100,000, it tells me that most employees make that amount. The max and min are certainly outliers because why does the most make $250,258 by the minimum is $0? That doesn't make a ton of sense. 

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

I think my story is why are there 11 white house employees that have a salary of $0. The next step I would do is to look up the names of each employee making $0 to see if they're some type of intern or if there's a reason for it.
