---
title: "HW4 MD CityCrime"
---
Benjamin Messinger

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
install.packages("tidyverse")
install.packages("janitor")
library(tidyverse)
library(janitor)

```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities |>
  mutate(
violent_rate_2019 = (violent_crime_2019 / population_2019) * 100000,
property_rate_2019 = (property_crime_2019 / population_2019) * 100000,
violent_rate_2020 = (violent_crime_2020 / population_2020) * 100000,
property_rate_2020 = (property_crime_2020 / population_2020) * 100000
)
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

The highest amount of crime in Maryland cities occurred in Baltimore through violent and property crimes in both years. Yes, Baltimore had the highest rate of crime because of it's large population, but other places were very close to as much. I definitely see the pattern that crime was less in 2020 in both violent and property crimes than in 2019. The existence of Covid in 2020, may have something to do with that. The story to me is how covid decreased the crime rate in Baltimore during 2020.

### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}

md_cities_change <- md_cities_rates |>
  mutate(
    violent_change_pct = ((violent_rate_2020 - violent_rate_2019) / violent_rate_2019) * 100,
    property_change_pct = ((property_rate_2020 - property_rate_2019) / property_rate_2019) * 100
  )

```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?


The largest violent percentage change was in Edmonston, it was extremely high in 2020 compared to 2019, I wonder why that is the case because of the nature of 2020. Cecilton had the highest property crime percentage change , but an extremely negative violent percentage change. Was there something that happened in Cecilton that caused this vast difference. Baltimore was negative in both. The story is that the percentage change for both property and violent crimes were negative. Meaning that crime as a whole was down.

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Crime rates significantly decreased in 2020 in Maryland most likely due to the Covid-19 pandemic.

In cities like Baltimore that usually see the highest crime rates in the state, saw both property and violent crime rates decrease from 2019 to 2020.

Despite some outliers that saw major jumps from 2019 to 2020, Maryland as a whole saw violent and property crime rates decrease consistently.

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

Hi, so my computer broke and I’m currently using a Chromebook that I’m not able to download R on, so I had to use AI in some of the code for this. I tried my best to use R on the computers at Knight Hall, and some of it worked, but it took a long time and wasn't loading properly. I apologize for the inconvenience. I did my best with the cards I’ve been dealt to complete this assignment, and I was able to review the data tables and calculate what was needed. 

