---
title: "HW6_Spotify_Songs"
---
Benjamin Messinger

In this assignment, you will be working with a dataset of songs played on Spotify in 2020, and you'll be writing code to explore the range of danceability scores in the data, using standard deviation and histograms. You'll also write up your findings.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)

```

### Task 2: Load the Spotify Songs dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv

```{r}
spotify_songs <- read.csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv")
```

### Task 3: Calculate the mean and standard deviation of the danceability scores

Write code to calculate the mean and standard deviation of the danceability scores in the Spotify dataset. Display those.

```{r}
spotify_songs |>
  summarize(
   mean_danceability = mean(danceability, na.rm=TRUE),
   sd_danceability = sd(danceability,na.rm=TRUE))
```
Mean = 0.6730774	Standard Deviation = 0.1615482
### Task 4: Create a histogram of the danceability scores

Write code to create a histogram of the danceability scores in the Spotify dataset, based on the example we did in class.

```{r}
spotify_songs |>
 ggplot() + 
  geom_histogram(aes(x = danceability), binwidth = 1)
```

### Task 5: Add a line to the histogram for the mean danceability score

Take the same code and add a vertical line to the histogram for the mean danceability score, based on the example in class.

```{r}
spotify_songs |>
  ggplot() + 
  geom_histogram(aes(x = danceability), binwidth = 1) +
  geom_vline(aes(xintercept = mean(danceability)), color = "red", linetype = "dashed", size = 1) 
```


### Task 6: Add two lines to the histogram to show one standard deviation

Take the previous code and add two lines to the histogram for the 1st standard deviation below and above the mean score.

```{r}
spotify_songs |>
  ggplot() + 
  geom_histogram(aes(x = danceability), binwidth = 1) +
    geom_vline(aes(xintercept = mean(danceability)), color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(danceability) - sd(danceability)), color = "blue", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(danceability) + sd(danceability)), color = "blue", linetype = "dashed", size = 1)
```

### Task 7: Describe our results and pitch a story

Write a few sentences about the distribution of danceability scores in the Spotify dataset. Is it a normal distribution, or is it skewed in one direction? What does the mean and one standard distribution tell us about the songs in the dataset? Draft a short story pitch from this data.

The majority of danceability scores are skewed to the right, that is because, the majority of scores are 0.5 or higher. The mean tells us that most danceability scores are 0.1615482 more or less than the mean of 0.6730774. Despite some outliers that have greater danceability scores. The highest score is 0.975 and the lowest score is 0.0. 

The story pitch for me is most songs on spotify have relatively same levels of danceability, I would investigate how danceability is calculated and what the measures of it are based on.



### Task 8: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
