---
title: "HW7_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse
```{r}

library(tidyverse)
```


### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv

```{r}

mm_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```


### Task 3: Sample 5 rows of data from the dataset

```{r}

sample_5 <- mm_data |>
sample_n(5)

```
### Task 4: Calculate the mean age and standard deviation of the sample, using either the blue or white number of M&Ms
```{r}
sample_5 |> summarize(mean = mean(black), sd =sd(black))

```
Mean: 72.2	sd: 9.338094	
### Task 5: Calculate the mean age and standard deviation of the population, using the same measure as in Task 4
```{r}
mm_data |> summarize(mean = mean(black), sd =sd(black))

```
mean: 73.28	sd:15.74871



### Task 6: Compare the sample and population statistics

Compare the two sets of means and standard deviations. Are they similar?
The means are similar at 72.2 for the sample and 73.28 for the full population, but the standard deviation is very different with the sample being 9.338094, but the population being 15.74871.

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

It four of the five examples in the sample, there were more red M&Ms than black M&Ms. To me, that means that random sampling has some sort of probability to it. It isn't a coincidence that the majority of the samples had more red M&Ms than black M&Ms, there is more red M&Ms in total for the majority of the 100 samples in the population, causing the random sample to display that. The difference in standard deviations of black M&Ms show that the five samples were more similar than the entire population, yet the means of black M&Ms were similar. 


When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
